---
title: "Lab 7"
author: Leo Hecht
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```



In this lab we will practice working with raster data, in this case in the context of climate models. I have given you 4 sets of data:

1. Climate Model Data_Historic - this is a NetCDF file with output from a climate model. Data is monthly average air temperature for 1920-2005
2. Climate Model Data_Future - this is a NetCDF file with climate model output for the period 2006-2080
3. Observed Temp Data - this is gridded data based on weather station and satellite data. Data is monthly for 1991-2000
4. Population data - gridded counts of population for a number of years

We will use the the packages raster, rnaturalearth, sf, tidyr, ggplot2, lubridate, and ncdf4.

The first part of the lab will compare modeled and observed climate data for major cities around the world. The second part of the lab will combine the population data and future climate data to project future changes in global temperature. 

```{r}
library("raster")
library("rnaturalearth")
library("sf")
library("tidyr")
library("ggplot2")
library("lubridate")
library("ncdf4")

```

## Part 1

1a. Read in the historic climate model data as a raster brick(). Use "TREFHT" (temperature at reference height) in the varname argument.
```{r}

file_path <- "C:/Users/Leo Hecht/Documents/ESP 106/Data/Climate Model Data_Historic/b.e11.B20TRC5CNBDRD.f09_g16.002.cam.h0.TREFHT.192001-200512.nc"

histclim <- brick(file_path, varname = "TREFHT")


```

1b. Use extent() to see the latitute and longitude of the raster brick you created in a. Note that the longitude goes form 0 to 360 (ish) instead of the more typical -180 to 180. This will cause a problem later on so use the rotate() function to change the longitude coordinates. Use extent again on the rotated object to check the longitude goes from -180 to 180 (ish)

```{r}

extent(histclim)

histclim_rotated <- rotate(histclim)

extent(histclim_rotated)
  
```


2a. USe the ne_download() function to get a sf object of major cities ("populated_places"). Subset to get just the 10 most populous cities based on 2020 population (POP2020 column)

```{r}


maj_cities <- ne_download(type = "populated_places") 

class(maj_cities$POP2020) 

maj_cities_ordered <- maj_cities[order(-maj_cities$POP2020), ]

top10 <- maj_cities_ordered[1:10, ]



```

2b. Make a plot of the temperature data for January 1920 and overlay the 10 major cities. I've given you this plot as Plot1 on Canvas.

```{r}

plot(histclim_rotated[[1]])
plot(top10, col = "black", pch = 16, add = TRUE)
title(main = "Temperature January 1920")

```

2c. What about the plot gives you confidence this is actually showing temperature data from a January? What are the units of the temperature data?

**Answer: The temperatures in the Northern hemisphere are much lower than the southern hemisphere, depicting the winter climate in that part of the world. The temperature is in degrees Kelvin.**

3a. Read in the observed temperature data as a raster brick, using "tmp" for the varname argument

```{r}

file_path2 <- "C:/Users/Leo Hecht/Documents/ESP 106/Data/Observed Temp Data/cru_ts4.03.1991.2000.tmp.dat.nc"

temp_raster <- brick(file_path2, varname = "tmp")


```

3b. Note that this climate model data is for 1920-2005 but the observation data is only from 1991-2000. Use subset() from the raster package to subset the climate model data to just the years 1991-2000.
```{r}

years <- rep(1920:2005, each = 12)

which(years %in% 1991:2000)

subset_model <- raster::subset(histclim_rotated, 853:972)

subset_model

```
4. Use the extract() from the raster package to produce two data-frames, one with observed and one with modeled temperature values for each city. Change the units of the modeled data so they match the units of the observed data.

```{r}

modeled <- as.data.frame(raster::extract(subset_model, top10))
modeled <- modeled - 273

observed <- as.data.frame(raster::extract(temp_raster, top10))

```

We have to do a bit of data-wrangling to compare modeled and observed temperature data for each city.

5a. Add a column to both data-frames with the names of the cities using the NAME column from the city data frame

```{r}


modeled$Name <- top10$NAME

observed$Name <- top10$NAME


```

5b. Use pivot_longer() from the tidyr package to turn both data-frames into tidy data-frames, with one row for each unique city-month combination

```{r}

modeled_pivot <- pivot_longer(modeled, cols = 1:120, values_to = "simulated")

observed_pivot <- pivot_longer(observed, cols = 1:120, values_to = "observed")

```

5c. Notice that the modeled and observed rasters have used slightly different conventions for naming the months. You can see this in the "name" column of the two data frames you made in 5b. The model output uses the first of the month (e.g. 1991.02.01) whereas the observational data uses the middle of the month (e.g. 1991.01.16). This is a problem since we want to merge together the two data frames to compare observed and simulated data.

To merge the two data frames together, first we need to "chop off" the last two digits in the month ids in both data frames. One way to do this is to use the substr() function to return some subset of a character vector.

Add a new column to both your data frames from 5b named "yearmonth" that strips out the day from the second data frame column to just leave the year and month data.

```{r}

modeled_pivot$yearmonth <- substr(modeled_pivot$name, 1, 8)

observed_pivot$yearmonth <- substr(observed_pivot$name, 1, 8)


```

5d. Merge together the observed and modeled city data into a single data-frame. 

```{r}

merged <- merge(modeled_pivot, observed_pivot, by = c("yearmonth", "Name"))


```

5e. Make a plot showing observed vs modeled temperature for the 10 cities. Add a 1:1 line which showing the exact match between observed and modeled data, and two dashed lines showing +/- 2 degree error. I have given you the plot you are aiming for as Plot2 on Canvas

```{r}

ggplot() +
  
  geom_point(data = merged, aes(x = merged$observed, y = merged$simulated, col = merged$Name)) +
  
  geom_abline(slope = 1) +
  
  geom_abline(slope = 1, intercept = 2, linetype = 2) +
  
  geom_abline(slope = 1, intercept = -2, linetype = 2) +
  
  labs(title = "Simulated vs. Observed Temperatures, 1991-2000", x = "Monthly Observed Temperature 1991-2000 (degrees C)", y = "Monthly Simulated Temperature 1991-2000 (degrees C)") +
  guides(col = guide_legend(title="City"))
  


```

## Part 2

In the second part of the lab, we will use projections of future temperature change (until 2080) and a map of the distribution of population in 2020 to get global, population-weighted projected warming.

6. Read in the netCDF file with projected climate model temperature (in the "Climate Model Data_Future" directory) as a raster brick. Use the rotate() function again as you did in 1b to transform the coordinates to -180 to 180. Use varname="TREFHT". This has gridded projections of monthly global temperature between 2006 and 2080 under a high-emissions scenario (referred to as RCP8.5).

```{r}

file_path <- "C:/Users/Leo Hecht/Documents/ESP 106/Data/Climate Model Data_Future/b.e11.BRCP85C5CNBDRD.f09_g16.002.cam.h0.TREFHT.200601-208012.nc"

future <- brick(file_path, varname = "TREFHT")

future_rotated <- rotate(future)

```

7a. Read in the netCDF data on population in the "Population" directory as a raster brick. (There is only one variable in this netCDF, so you can leave the varname argument empty this time). This is gridded population count at 15 arc minute resolution.

```{r}

file_path <- "C:/Users/Leo Hecht/Documents/ESP 106/Data/Population/gpw_v4_population_count_adjusted_rev11_15_min.nc"

population <- brick(file_path)

```

7b. We want only the 5th layer in this raster brick, which corresponds to population count in 2020. (Note - I know this from some associated files that came with the netCDF file. Take a look at the csv file in the directory to see this documentation). Use the subset() function from the raster package to pull out just the population in 2020.

```{r}

population_2020 <-  raster::subset(population, 5)

population_2020

```

8a. Now we want to eventually match the population grid to the projected temperature grid. But the problem is that the grid size of the climate model is much larger than the grid size of the population data. How many rows and columns does the climate model data have? How many rows and columns does the population data have?

```{R}

future_row <- nrow(future_rotated)
future_col <- ncol(future_rotated)

pop_row <- nrow(population_2020)
pop_col <- ncol(population_2020)

```

**Answer: The climate model data has 192 rows and 288 columns. The population data has 720 rows and 1440 columns.**

8b. To fix this problem we can aggregate the population raster up to the resolution of the climate model using the aggregate() function in the raster package. The population data you have is the population count (i.e. number of people in each grid cell). What function should we use to aggregate to larger grid cells? What function would we use instead if we had population density data instead of population count?

**Answer: We should use sum. If we had population density, we would need to take the mean.**

8c. Aggregate the population data to a higher level of resolution, as close as possible to the climate model data.

```{r}

row_factor <- (pop_row / future_row)

col_factor <- (pop_col / future_col)


pop_2020_aggregated <- raster::aggregate(population_2020, fact = c(col_factor, row_factor), fun = sum)


ncol(pop_2020_aggregated)

nrow(pop_2020_aggregated)


```

8d. If everything has gone according to plan, we would expect that summing up all the cells in the population raster should give us something close to the current population on the planet. Calculate that sum from your aggregated population data and compare to the total population in 2020.

```{r}

total_pop_2020 <- cellStats(pop_2020_aggregated, stat = "sum")

total_pop_2020

```

**Answer: My aggregated population is 7,757,982,600 which is close to the real 2020 world population of 7.8 billion.**

9a. Now we will use the population data to do a weighted averaging of the projected temperature data, to get the monthly temperature experienced by the average person between 2006 and 2080. 

One problem is that even after the aggregation, the grids of the population data still don't quite match. Use the resample() function in the raster package to resample the aggregated population data to the climate model grid. 

```{r}

pop_2020_resampled <- resample(pop_2020_aggregated, future_rotated)

pop_2020_resampled

```

9b. Now we can use the population raster to do a weighted average of the climate model data. Use the cellStats() function to calculate the population-weighted average temperature for each month. 

Convert from the units of the climate model data into degrees C

```{r}

temp <- future_rotated * pop_2020_resampled

weighted_average <- cellStats(temp, stat = sum) / cellStats(pop_2020_resampled, stat = sum)

weighted_average_C <- weighted_average - 273

head(weighted_average_C)


```

9c. Run the code below to get the dates of each month from the netCDF file of future climate model data, substituting in the location of the climate model projections on your computer where I have put "FILENAME". You will need to install the ncdf4 package. 

This is an example of how to access netCDF data other than through the raster package. You first open a connection to the file (nc_open()) and then read in just the data you want (ncvar_get()).

```{r}
library(ncdf4)
library(lubridate)

file=nc_open("C:/Users/Leo Hecht/Documents/ESP 106/Data/Climate Model Data_Future/b.e11.BRCP85C5CNBDRD.f09_g16.002.cam.h0.TREFHT.200601-208012.nc") 
date=ncvar_get(file,"date")
date=ymd(date)

```

9d. Create a data-frame with two columns - the population-weighted temperature data (from 9b) and the date (from 9c)

```{r}

weighted_pop_dated <- data.frame(weighted_average_C, date)

head(weighted_pop_dated)


```

10a. Make a plot of population-weighted global temperature. Add a smoothing line to show the average trend. The plot you are aiming for is on Canvas as Plot3.

```{r}


ggplot(data = weighted_pop_dated) +
  
  geom_line(aes(x = date, y = weighted_average_C)) +
  
  geom_smooth(aes(x = date, y = weighted_average_C), method = lm, se = TRUE) +
  
  labs(title = "Population-Weighted Global Temperature (RCP 8.5)", x = "Date", y = "Temperature (degrees C)")



```

10b. How do you think the rate of warming might be different if we were to calculate an area-weighted average across the globe, instead of the population weighted average shown in the plot from 10a? 

**Answer: The difference in the rate of warming in area-weighted average would an a population weighted average would depend on if warming is accelerating more or less in population centers than the average global rate of increase. If warming is more significant in areas of high population, the slope would be steeper. If warming is less significant in areas of high population, the slope would be shallower.**