---
title: "Lab 6" 
author: Leo Hecht
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

In this lab we will use some GPS tracking data for Blue Whales in the eastern North Pacific, downloaded from Movebank <https://www.movebank.org> which has lots of wildlife tracking data. We will make some cool maps and overlay polygons of Marine Protected Areas off the California coast to understand how whales move into and out of MPAs. 

We will need the ggplot2, sf, rnaturalearth, rnaturalearthdata and viridis packages

<br />
1. Use st_read to read in the datapoints for the Blue Whale migration data (poinsts.shp file). This is a set of GPS location points for blue whales tagged in the eastern north Pacific. Identifiers for individual whales are given in the ind_ident column. How many individual whales are tracked in this dataset?

```{r}


library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)

```
  

```{r}

whale_migration <- st_read("C:\\Users\\Leo Hecht\\Documents\\ESP 106\\Blue Whale Migration Data\\points.shp")

num_whales <- length((unique(whale_migration$ind_ident)))

num_whales



```
<br />
2a. The package rnaturalearth and rnaturalearthdata have lots of useful spatial data and functions for interacting with it. Load these packages and use the ne_countries() function to get a country map of north america (contient="North America") at medium resolution (scale argument) as an sf object  

```{r}


NA_map <- ne_countries(scale = "medium", continent = "North America", returnclass = "sf")

head(NA_map)

```
<br />
2b. Use the identical() function to test whether the coordinate reference system (crs) is the same for the GPS location points and the north america map. If it is not, use st_transform() to transform the north america map to the same crs as the GPS points. If necessary, use the identical() function again to check that the crs of the two data-sets are the same.
  
```{r}

identical(st_crs(whale_migration), st_crs(NA_map))


```
<br />
3. Use ggplot2 to make a plot showing GPS location points for each whale in a different color. Add the west coast of North America for reference. Set the limits of the plot to the extent of the whale migration data. Plot 1 on canvas shows you the plot you are aiming for here.  

```{r}

st_bbox(whale_migration)



ggplot() +
  
  geom_point(data = whale_migration, aes(x = long, y = lat, col = ind_ident), show.legend = FALSE) +
  
  geom_sf(data = NA_map, inherit.aes = FALSE) +
  
  coord_sf(xlim = c(-151.570, -87.547), c(4.606, 53.148)) +
  
  labs(title = "Blue Whale Migration Data", x = NULL, y = NULL)




```

<br />
4. The folder MPAs has a shapefile in it with polygons for Marine Protected Areas within the United States. I created this file based on polygons for all protected areas in the US, which came from <https://www.protectedplanet.net>. Read in the MPA shapefile and transform the crs to match that of the whale dataset.  

```{r}

MPAs <- st_read("C:\\Users\\Leo Hecht\\Documents\\ESP 106\\MPAs\\mpas.shp")

MPAs <- st_transform(MPAs, st_crs(whale_migration))



```
<br />
5. Now we will zoom into just the US west coast. You can use the following lognitude and latitude coordinates: xmin=-126.5, xmax=-116, ymin=32, ymax=40 as a new bounding box. Get just the MPAs on the US west coast by cropping the MPA dataset to this bounding box  


```{r}

sf_use_s2(FALSE)

MPAs = st_make_valid(MPAs)

cropped <- c(xmin=-126.5, ymin=32, xmax=-116, ymax=40)

MPAs_cropped <- st_crop(MPAs, cropped)

head(MPAs_cropped)

NA_map_cropped <- st_crop(NA_map, cropped)

  
```
<br />
6. Find what fraction of blue whales in the dataset spend at least some time in one of these west coast MPAs  


```{r}

intersects <- st_intersection(unique(whale_migration), MPAs_cropped) 

ratio <- nrow(intersects) / nrow(unique(whale_migration))

ratio


```
<br />
7. Now we will make a map of blue whale density along the California coast and overlay the west coast MPAs. Use ggplot to create another map of the blue whale data, but crop to the smaller extent (given in question 5). Rather than plotting the point data, use stat_density_2d() to plot a smoothed out, continuous map of whale locations. Add the north american contient and the west coast MPA polygons. Plot 2 on Canvas shows you the plot you are aiming for.  


```{r}



ggplot() +
  
  stat_density2d(data = whale_migration, aes(x = long, y = lat, fill = after_stat(density)), geom = "raster", contour = FALSE, n = 175) +
  
  geom_sf(data = MPAs_cropped, aes(), fill = "transparent", color = "black") +
  
  scale_fill_viridis() +
  
  geom_sf(data = NA_map_cropped) +
  
  coord_sf(xlim = c(-126.5, -116), ylim = c(32, 40), expand = FALSE) +
  
  labs(title = "Blue Whale Migration Data and Marine Protected Areas", x = NULL, y = NULL ) +
  
  guides(fill = guide_legend(title = "Density")) + 
  
  theme(plot.title = element_text(size = 12.2))


 


```