---
title: "Lab 5"
author: Leo Hecht
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:\\Users\\Leo Hecht\\Documents\\ESP 106") #use this to set your working directory so knitr understands it
```


Due Tuesday Feb1st - Recommended to complete this before starting the midterm

This lab we will look at some data from the plastic trash picked up during clean-up events around the world. I took this dataset from the Tidy Tuesday website. You can read the documentation [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-01-26/readme.md), including the references and description of the different column names.

I have done some pre-processing of the data for you for this lab, to create two more easy-to-use dataframes.

First read in the countrytotals.csv data frame
```{r}

countrytotals <- read.csv("countrytotals.csv")

```

Have a look at the data frame. Then column "total" gives the total number of pieces of plastic picked up in that country in 2020. The columns "num_events" and "volunteers" give the number of trash pick-up events and the number of volunteers in that country. We are going to use this to investigate where the plastic trash problem is worst.

## 1. What 5 countries had the worst plastic problem as measured by the number of pieces of trash picked up?  

```{r}


countrytotals <- countrytotals[order(countrytotals$total, decreasing = TRUE), ]

countrytotals[1:5, 2]


```

**Answer: Nigeria, Philippines, Switzerland, India, and Togo**  

## 2. Make a plot showing the distribution of volunteers across countries  

```{r}

library(ggplot2)

ggplot(countrytotals, aes(country, volunteers)) +
  geom_bar(stat = "identity")


```

## 3. Notice that there is a lot of variation across countries in the number of volunteers involved in trash pickup. What problem might that cause for the interpretation of your answer to question 1?  

**Answer: Countries with many volunteers could have artificially inflated totals of plastic as there were more people to pick it up. Conversely, there could be countries with high pollution that lack volunteers, meaning the plastic pollution was under counted.**  

## 4. Add a column to the data frame creating a variable that should be more closely related to the presence of plastic pollution in the country   

```{r}

countrytotals$pollution_per_volunteer <- countrytotals$total / countrytotals$volunteers


```

## 5. What 5 countries have the worst plastic pollution, as measured by this new variable?  

```{r}
countrytotals <- countrytotals[order(countrytotals$pollution_per_volunteer, decreasing = TRUE), ]

countrytotals[1:5, 2]

```

**Answer: Togo, El Salvador, Japan, Kuwait, and Germany** 

Now we will make a plot of the variation in the types of trash and how it differs around the world. Read in the continenttypes.csv data frame. This gives the breakdown of the different types of plastic collected on each continent in 2020 and the total number of pick up events.

```{r}

continent_types <- read.csv("continenttypes.csv")



```

## 6. Add a column to this data frame with a variable that captures the existence of different types of plastic trash, controlling for the intensity of the pick-up effort in different continent  

```{r}

continent_types$types_controlled <- continent_types$total / continent_types$events


```

## 7. Make a plot using ggplot showing both the total amount and distribution of types of plastic picked up in each continent in the average pick-up event.  


```{r}
ggplot(continent_types, aes(fill = plastic_type, x = continent, y = types_controlled)) +
  geom_bar(position="stack", stat="identity") +
  labs(x = "Continent", y = "Trash Recovered Per Event (pieces)", fill = "Plastic Type")


```

## 8. Try uploading your R markdown file and plot to your Git Hub repository. Don't put your knitted HTML file in Github - these are large files that are not designed to be stored on Github

